Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-30 manuscript/manuscript.Rmd
first complete draft (with holes) and render

#Want to minimize false negatives.  Ratio of FN/FP+TP+TN

#mc_chla[7] <- get_chla(who_rec_med2_cp,0.5)
```

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down their municipal water supply due in part to an excess of microcystin-LR that resulted from a ongoing cyanobacteria harmful algal bloom (cHAB) in Lake Erie [@jetoo2015toledo; @rinta2009lake].  Since this event, significant legislation has been passed in the United States and the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic].  MORE ON THE LEVELS. While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

@@

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives [REF].  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`, `r fn_percent$who_drink*100`, `r fn_percent$epa_adult*100` and , `r fn_percent$who_rec*100 for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational advisories, respecitvely.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.  This is an acceptable rate, especially when no other tests are available for a given water body.  TEST LOWER VALUES to see where we can get it even lower.

@@

These values are conservative: greater than even odds for the upper CI. Most protective.

Not meant as a replacement for testing of microcystin, but provides another useful tool to help understand the probable extent of microcystin occurence.

>>>>

```

@@

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; NEED HABHRCA REFS].  Recent, the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [REF].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

@@

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives [REF].  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`, `r fn_percent$who_drink*100`, `r fn_percent$epa_adult*100` and , `r fn_percent$who_rec*100 for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational advisories, respecitvely.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.  This is an acceptable rate, especially when no other tests are available for a given water body.

@@

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous meausrement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases.

### 2015-09-30 manuscript/manuscript.tex
first complete draft (with holes) and render

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and require advanced understanding to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 22.56, 59.4, 84.24, and 97.49, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

In the summer of 2014, the city of Toledo, OH was forced to shut down
their municipal water supply due in part to an excess of microcystin-LR
that resulted from a ongoing cyanobacteria harmful algal bloom (cHAB) in
Lake Erie (Rinta-Kanto et al. 2009, Jetoo et al. 2015). Since this
event, significant legislation has been passed in the United States and
the US Environmental Protection Agency (USEPA) has released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005). MORE ON THE LEVELS. While these levels and associated advisories
are likely to help mitigate the impacts from harmful algal blooms, they
are not without complications.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 0.51. Lastly, of all lakes
sampled, 21\% of lakes exceeded the U.S. EPA childrens drinking water
standard, 8.8\% of lakes exceeded the U.S. EPA adult drinking water
standard,11.7\% of lakes exceeded the WHO drinking water standard,and
7.3\% of lakes exceeded the WHO recreational standard. For chlorophyll
\emph{a}, the range was 0.07 to 936. All lakes had reported chlorophyll
\emph{a} concentrations that exceeded detection limits. The median
concentration was 7.79 and the mean was 29.6301946. The associations
between chlorophyll \emph{a} and the upper confidence interval with a
conditional probability of 50\% were 22.56, 59.4, 84.24, and 97.49 for
0.3, 1.0, 1.6 and 2.0 \(\mu\)g/L, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 28.8\%, 12.7\%, 8.8\%, and
7.9\% of sample lakes for the U.S EPA childrens drinking water, the WHO
drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational advisories, respecitvely.

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 74\% for the USEPA children's advisory,
85\% for the WHO drinking water advisory, 89\% for the USEPA adult
advisory and, 90\% for the WHO regreational advisory. But Total Accuracy
is one part of the prediction performace with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives {[}REF{]}. For the four chlorophyll \emph{a} cut-offs
we had a proportion of false negatives of 9, 7, 6 and , `r
fn\_percent\$who\_rec*100 for the U.S EPA childrens drinking water, the
WHO drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational advisories, respecitvely. In each case we miss less than
10\% of the lakes that are in fact exceeding the microcystin-LR
advisory. This is an acceptable rate, especially when no other tests are
available for a given water body. TEST LOWER VALUES to see where we can
get it even lower.

@@

These values are conservative: greater than even odds for the upper CI.
Most protective.

Not meant as a replacement for testing of microcystin, but provides
another useful tool to help understand the probable extent of
microcystin occurence.

@@

\begin{longtable}[c]{@{}llrr@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin-LR health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endfirsthead
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endhead
U.S. EPA & Drinking & 0.3 & 22.56\tabularnewline
WHO & Drinking & 1.0 & 59.40\tabularnewline
U.S. EPA & Drinking & 1.6 & 84.24\tabularnewline
WHO & Recreational & 2.0 & 97.49\tabularnewline
\bottomrule
\end{longtable}

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and require advanced understanding to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 22.25, 65.6, 84.96, and 105.84, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms {[}Jetoo et al.
(2015); Rinta-Kanto et al. (2009); NEED HABHRCA REFS{]}. Recent, the US
Environmental Protection Agency (USEPA) has released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005). Additionally, the World Health Organization (WHO) has had
proposed advisory levels for drinking water and a range of recreational
risk levels {[}REF{]}. While these levels and associated advisories are
likely to help mitigate the impacts from harmful algal blooms, they are
not without complications.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 0.51. Lastly, of all lakes
sampled, 21\% of lakes exceeded the U.S. EPA childrens drinking water
standard, 8.8\% of lakes exceeded the U.S. EPA adult drinking water
standard,11.7\% of lakes exceeded the WHO drinking water standard,and
7.3\% of lakes exceeded the WHO recreational standard. For chlorophyll
\emph{a}, the range was 0.07 to 936. All lakes had reported chlorophyll
\emph{a} concentrations that exceeded detection limits. The median
concentration was 7.79 and the mean was 29.6301946. The associations
between chlorophyll \emph{a} and the upper confidence interval with a
conditional probability of 50\% were 22.25, 65.6, 84.96, and 105.84 for
0.3, 1.0, 1.6 and 2.0 \(\mu\)g/L, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 29.1\%, 11.5\%, 8.6\%, and
6.9\% of sample lakes for the U.S EPA childrens drinking water, the WHO
drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational advisories, respecitvely.

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 74\% for the USEPA children's advisory,
86\% for the WHO drinking water advisory, 90\% for the USEPA adult
advisory and, 91\% for the WHO regreational advisory. But Total Accuracy
is one part of the prediction performace with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives {[}REF{]}. For the four chlorophyll \emph{a} cut-offs
we had a proportion of false negatives of 9, 7, 6 and , `r
fn\_percent\$who\_rec*100 for the U.S EPA childrens drinking water, the
WHO drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational advisories, respecitvely. In each case we miss less than
10\% of the lakes that are in fact exceeding the microcystin-LR
advisory. This is an acceptable rate, especially when no other tests are
available for a given water body.

@@

Lastly, using chlorophyll \emph{a} is not meant as a replacement for
testing of microcystin-LR. It should be used when other, direct
measurements of cyanotoxins are not available. In those cases, which are
likely to be common at least in the near future, using a more ubiquitous
meausrement, such as chlorophyll \emph{a} will provide a reasonable
proxy for the probability of exceeding a microcystin-LR health advisory
level and provide better protection against adverse effects in both
drinking and recreational use cases.

@@

\begin{longtable}[c]{@{}llrr@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin-LR health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endfirsthead
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endhead
U.S. EPA & Drinking & 0.3 & 22.25\tabularnewline
WHO & Drinking & 1.0 & 65.60\tabularnewline
U.S. EPA & Drinking & 1.6 & 84.96\tabularnewline
WHO & Recreational & 2.0 & 105.84\tabularnewline
\bottomrule
\end{longtable}
